Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
added Lemma in appendix B to explain better a passage

\begin{table*}[t]
	\caption{
		\label{tab:1}
		Comparison between G(PO)MDP and SVRPG on different tasks. Average return over the whole learning process is reported, together with normalized standard deviation and 90\% confidence intervals.}
	\centering
	\begin{tabular}{@{}lccccccc@{}} 
		\toprule
		\phantom{abc} & \multicolumn{3}{c}{G(PO)MDP} & \phantom{abc} & \multicolumn{3}{c}{SVRPG} \\
		\cmidrule{2-4} \cmidrule{6-8}
		\phantom{abc} & Avg Return & Norm. Std & C.I. & \phantom{abc}
		& Avg Return & Norm. Std & C.I.
		\\\cmidrule{2-8}
		Cart-pole & 618.49 & 21.09 & [579.82, 657.16] & \phantom{abc} 
		& \textbf{736.31} & 16.51 & [706.04, 766.57] \\ 
		\phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} \\
		Swimmer & 39.86 & 3.84 & [32.82, \:\:\: 46.89] & \phantom{abc} & 49.60 & 3.27 & [43.60, \:\:\: 55.60]\\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\begin{definition}
For random vectors X, Y:
\begin{align*}
	\Covs{X}{Y} &\coloneqq \Tr\left(\Ets{(X-\Ets{X})(Y-\Ets{Y})^T}\right), \\
	\Vars{X} &\coloneqq \Covs{X}{Y},
\end{align*}
where $\Tr(\cdot)$ denotes the trace of a matrix. From the linearity of expected value we have the following:
\begin{equation}\label{neweq:1}
\Vars{X} = \Es{\norm[]{X-\Es{X}}^2} \leq \Es{\norm[]{X}^2}
\end{equation}
$\Covts[t]{X}{Y}$ and $\Varts[t]{X}$ are defined in the same way from $\Ets[t]{X}$.
\end{definition}

@@

\begin{restatable}[]{lemma}{bounded-g}\label{lemma:gbound}
Under Assumption \ref{ass:bounded_score}, whether we use the REINFORCE or the G(PO)MDP gradient estimator, for every $\tau\in\Tspace$ and $\vtheta\in\Theta$, there is a positive constant $\Gamma<\infty$ such that:
	\norm[2]{g(\tau\vert\vtheta)}^2 \leq \Gamma.
For REINFORCE we have, from Assumption \ref{ass:bounded_score}:
	\norm[2]{g(\tau\vert\vtheta)}^2 &=
	\norm[2]{\score{\vtheta}{\tau}R(\tau)}^2 \\
	&=\norm[2]{\left(\sum_{t=0}^{H-1}\score{\vtheta}{a_t\vert s_t}\right)\left(\sum_{t=0}^{H-1}\gamma^t r_t\right)}^2\leq H^2G^2\frac{(1-\gamma^H)^2}{(1-\gamma)^2}R^2\dim(\vTheta) \coloneqq \Gamma
For G(PO)MDP, we do not have a compact expression for $g$, but since it is derived from REINFORCE by neglecting some terms of the kind $\nabla \log \pi_{\vtheta}{a_t | s_t}\gamma^t r_t$ \cite{baxter2001infinite,peters2008reinforcement}, the above bound still holds.

@@

where (\ref{neweq:2}) is from (\ref{neweq:1}), (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from (\ref{neweq:1}), (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
Under Assumption \ref{ass:bounded_score}, for any function $\varphi(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$:
\begin{align*}
\left|\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
-\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
\right|
\leq
\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:6}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:7}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{\varphi(\vtheta_t^{s+1})}}
	+\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}  \label{eq:new1}\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})} \label{eq:8}
\end{align}
where~\eqref{eq:6} is from Definition~\ref{def:ideal};~\eqref{eq:7} is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $\varphi(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$;~\eqref{eq:new1} is from the fact that $\gradJ{\wt{\vtheta}^s}$ is constant \wrt $\Vars{\cdot}$;~\eqref{eq:8} is from $\Ets{e_s}=0$.
Hence:
\begin{align}
	\left|\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	\right.&-\left.\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}\right| 
	=
	\left|\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}\right|  
	\nonumber\\
	&\leq
    \sqrt{\Vars{\gradApp{\wt{\vtheta}^s}{N}}} \cdot \sqrt{\Varts[t-1]{\varphi(\vtheta_t^{s+1})}} \label{eq:9a}\\
	%
	&\leq	
	\frac{1}{2}\Vars{\gradApp{\wt{\vtheta}^s}{N}} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})}\label{eq:9}\\
	%
	&=
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})} \label{eq:10}\\
	%
	&\leq
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2},
	\label{neweq:3}
\end{align}
where~\eqref{eq:9a} comes from Cauchy-Schwarz inequality
\[
        |\Cov(X,Y)| = |\EVV[]{(X-\mu_X)\transpose{(Y-\mu_Y)}}| \leq \EVV[]{(X-\mu_X)^2}^{1/2}\EVV[]{(Y-\mu_Y)^2}^{1/2} = \sqrt{\Var(X) \Var(Y)},
\]
\eqref{eq:9} is from Young's inequality,~\eqref{eq:10} is from the definition of $\gradApp{\vtheta}{N}$, and \eqref{neweq:3} is from \eqref{neweq:1}.
\end{proof}

>>>>

\begin{table*}[t]
	\caption{
		\label{tab:1}
		Comparison between G(PO)MDP and SVRPG on different tasks. Average return over the whole learning process is reported, together with normalized standard deviation and 90\% confidence intervals.}
	\centering
	\begin{tabular}{@{}lccccccc@{}} 
		\toprule
		\phantom{abc} & \multicolumn{3}{c}{G(PO)MDP} & \phantom{abc} & \multicolumn{3}{c}{SVRPG} \\
		\cmidrule{2-4} \cmidrule{6-8}
		\phantom{abc} & Avg Return & Norm. Std & C.I. & \phantom{abc}
		& Avg Return & Norm. Std & C.I.
		\\\cmidrule{2-8}
		Cart-pole & 618.49 & 21.09 & [570.76, 666.20] & \phantom{abc} 
		& \textbf{736.31} & 16.51 & [698.96, 773.65] \\ 
		\phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} \\
		Swimmer & 39.86 & 3.84 & [31.18, \:\:\: 48.53] & \phantom{abc} & 43.33 & 3.27 & [37.41, \:\:\: 49.25]\\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\begin{definition}\label{def:var}
For random vectors X, Y:
\begin{align*}
	\Covs{X}{Y} &\coloneqq \Tr\left(\Ets{(X-\Ets{X})(Y-\Ets{Y})^T}\right), \\
	\Vars{X} &\coloneqq \Covs{X}{Y},
\end{align*}
where $\Tr(\cdot)$ denotes the trace of a matrix. From the linearity of expected value we have the following:
\begin{equation}\label{neweq:1}
\Vars{X} = \Es{\norm[]{X-\Es{X}}^2}
\end{equation}
$\Covts[t]{X}{Y}$ and $\Varts[t]{X}$ are defined in the same way from $\Ets[t]{X}$.
\end{definition}

@@

\begin{restatable}[]{lemma}{bounded-g}\label{lemma:gbound}
Under Assumption \ref{ass:bounded_score}, whether we use the REINFORCE or the G(PO)MDP gradient estimator, for every $\tau\in\Tspace$ and $\vtheta\in\Theta$, there is a positive constant $\Gamma<\infty$ such that:
\[
	\norm[2]{g(\tau\vert\vtheta)}^2 \leq \Gamma.
\]
\end{restatable}
\begin{proof}
For REINFORCE we have, from Assumption \ref{ass:bounded_score}:
\begin{align*}
	\norm[2]{g(\tau\vert\vtheta)}^2 &=
	\norm[2]{\score{\vtheta}{\tau}R(\tau)}^2 \\
	&=\norm[2]{\left(\sum_{t=0}^{H-1}\score{\vtheta}{a_t\vert s_t}\right)\left(\sum_{t=0}^{H-1}\gamma^t r_t\right)}^2\leq H^2G^2\frac{(1-\gamma^H)^2}{(1-\gamma)^2}R^2\dim(\vTheta) \coloneqq \Gamma
\end{align*}
For G(PO)MDP, we do not have a compact expression for $g$, but since it is derived from REINFORCE by neglecting some terms of the kind $\nabla \log \pi_{\vtheta}{a_t | s_t}\gamma^t r_t$ \cite{baxter2001infinite,peters2008reinforcement}, the above bound still holds.
\end{proof}

%New Lemma on Variance
\begin{restatable}[]{lemma}{varineq}\label{lemma:varineq}
For any random vector X, the variance (as defined in Definition \ref{def:var}), can be bounded as follows:
\[
	\Vars{X} \leq \Es{\norm[]{X}^2}.
\]
\end{restatable}
\begin{proof}
By using basic properties of expected value and scalar variance:
\begin{align*}
	\Vars{X} = \Es{\norm[]{X-\Es{X}}^2} &= \Es{\sum_{i=1}^{\dim(X)}\left(X_i-\Es{X_i}\right)^2} = \sum_{i=1}^{\dim(X)}\Es{\left(X_i-\Es{X_i}\right)^2} \\
	&\leq \sum_{i=1}^{\dim(X)}\Es{X_i^2} = \Es{\sum_{i=1}^{\dim(X)}X_i^2} = \Es{\norm[]{X}^2}.
\end{align*}
\end{proof}

@@

where (\ref{neweq:2}) is from (\ref{neweq:1}), (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from Lemma \ref{lemma:varineq}, (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
Under Assumption \ref{ass:bounded_score}, for any function $\varphi(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$:
\begin{align*}
\left|\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
-\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
\right|
\leq
\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:6}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:7}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{\varphi(\vtheta_t^{s+1})}}
	+\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}  \label{eq:new1}\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})} \label{eq:8}
\end{align}
where~\eqref{eq:6} is from Definition~\ref{def:ideal};~\eqref{eq:7} is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $\varphi(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$;~\eqref{eq:new1} is from the fact that $\gradJ{\wt{\vtheta}^s}$ is constant \wrt $\Vars{\cdot}$;~\eqref{eq:8} is from $\Ets{e_s}=0$.
Hence:
\begin{align}
	\left|\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	\right.&-\left.\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}\right| 
	=
	\left|\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}\right|  
	\nonumber\\
	&\leq
    \sqrt{\Vars{\gradApp{\wt{\vtheta}^s}{N}}} \cdot \sqrt{\Varts[t-1]{\varphi(\vtheta_t^{s+1})}} \label{eq:9a}\\
	%
	&\leq	
	\frac{1}{2}\Vars{\gradApp{\wt{\vtheta}^s}{N}} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})}\label{eq:9}\\
	%
	&=
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})} \label{eq:10}\\
	%
	&\leq
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2},
	\label{neweq:3}
\end{align}
where~\eqref{eq:9a} comes from Cauchy-Schwarz inequality
\[
        |\Cov(X,Y)| = |\EVV[]{(X-\mu_X)\transpose{(Y-\mu_Y)}}| \leq \EVV[]{(X-\mu_X)^2}^{1/2}\EVV[]{(Y-\mu_Y)^2}^{1/2} = \sqrt{\Var(X) \Var(Y)},
\]
\eqref{eq:9} is from Young's inequality,~\eqref{eq:10} is from the definition of $\gradApp{\vtheta}{N}$, and \eqref{neweq:3} is from Lemma \ref{lemma:varineq}.
\end{proof}
